Synthesizing Symmetric Lenses

### 2018-07-12 papers/SymmetricLenses/main.tex
change alg to 60 sec
add more info

One convenient way to interconvert between these varied data formats is to
use a \emph{lens}~\cite{Focal2005}---a bidirectional program that can
transform data represented in a format $S$ to a format $T$, and vice versa.
offering ``round tripping'' guarantees that help ensure data is not lost or
corrupted as it is transformed back and forth between different
representations.

While domain-specific languages for writing lenses can facilitate building
principled data transformations, \textit{synthesizing} lenses 
can make the task even easier.  Indeed, lens languages offer excellent
targets for synthesis, since their specialized, sub-Turing-complete syntax
and expressive type systems drastically
reduce the space of possible programs, making search-based synthesis easier. 
%
Recent work~\cite{optician} has exploited this
observation, demonstrating that it is feasible to synthesize quite complex
examples, albeit only for a simple form of lenses---so-called bijective
ones. Specifically, given a pair of types describing the intended
source and target formats plus a small number of examples of corresponding
data, the bijective synthesis algorithm developed in these two
papers will find a bijective lens that converts between data stored
in different formats,
such as linux configuration files from different systems that represent the same data in different ways.  
Often, these algorithms can find an appropriate lens
in less than a second without requiring any examples.
%, with just types describing the source and target formats.
% \dpw{I would cite the ICFP paper (cite string:
%   maina+:quotient-synthesis) as to appear but I'm not sure how to do that
%   without possibly violating the spirit of double-blind reviewing. ie, if it
%   hasn't appeared yet, how could we know the content if we aren't the authors?
%   Mentioning it in the 3rd person does not seem to suffice ...} \ksf{The ICFP
%   2018 website lists the paper; I think it is okay to cite it. The reviewer can
%   assume we contacted the authors for a copy. We could ask Stephanie for a
%   ruling if we want to be sure. One of the principles of double blind as adopted
%   by SIGPLAN is not to reduce the quality of the paper; omitting this citation
%   would have that affect.}\bcp{+1}\saz{Stephanie says we should cite it in the
%   third person, even if that's a bit awkward, but otherwise citing it is OK.}
One reason that bijective lens synthesis is so effective, even relative to successful
synthesis projects in other domains, is that there are not very
many bijections between related data formats (unless the
same kind of data appears in the two formats many times). 
Hence, if the synthesis algorithm can
find any bijection at all, then it is very likely to find the one the
user wants.

However, there is a price for narrowing the synthesis domain so
dramatically: Many real data formats are almost, but not quite,
bijections. More often two related data formats have some data in
common, but also contain some differences. For instance, two related
tables might share four of five columns, but each has a different
fifth column. Alternately, one ad hoc configuration file might include
some file-specific meta data, such as a date, a time or a reference
number, whereas the other does not. Indeed of the 48 benchmark problems
described in Section \ref{sec:evaluation}, all of the benchmarks
taken from Flash Fill~\cite{flashfill} and many of the benchmarks that
synchronize between two ad hoc file formats have this 
characteristic, 13 benchmarks in total. Consequently, tools that search
only for strict bijections are often not applicable in practice.

@@

Initially, we hypothesized that extending the synthesis algorithms
of Miltner \ETAL~\cite{optician} to operate over symmetric rather than
bijective lenses
would be relatively straightforward:  Simply replace the bijective
combinators with symmetric ones 
and search using similar heuristics.  However, 
this na\"ive approach runs into two difficulties.

@@

In summary, our contributions are:\bcp{The section references do not proceed
  in order, so this doesn't work very well as an outline of the paper.
  Should we try to reorder the contributions?  Write a separate outline?
  Live with it as-is?}\afm{Personally, I don't see it as a problem.  It's nice
  when the contributions mimic the paper, I am not jarred when they don't.}
\begin{itemize}
\item We define {\em simple symmetric lenses}, with the
  the useful property that they do not maintain any hidden state
  (\S\ref{sec:overview}).  
  We present a collection of combinators for defining these lenses and
  describe how to 
  calculate their likelihoods (\S\ref{sec:ssl}). 
\item  We show that simple symmetric lenses
  fall between asymmetric lenses~\cite{Focal2005-long2} and full
  symmetric lenses~\cite{symmetric-lenses} in expressiveness
  (\S\ref{sec:related}).  We further show that the class of simple
  lenses can be characterized semantically as the subset of full
  symmetric lenses for which a simple ``forgetfulness'' property holds.
\item We show how to synthesize simple symmetric lenses (including all
  asymmetric lenses as a special case), using a novel application of stochastic
  regular expressions to guide the search process (\S\ref{sec:synthesis}). 
  To support that effort, we show that the star semi-ring equivalences on
  regular 
  expressions remain valid when applied to stochastic regular
  expressions, preserving the probability distributions as well as the languages
  (\S\ref{sec:sre}).
\item We extend the Boomerang implementation with simple symmetric lenses,
  demonstrating that they integrate smoothly with 
  other lens extensions such as quotient lenses~\cite{quotientlenses} and matching
  lenses~\cite{matchinglenses}.  We evaluate our
  implementation on a set of 48 lens synthesis benchmarks drawn from
  data cleaning, view maintenance, 
  and file synchronization tasks.  We demonstrate that we can
  synthesize simple symmetric lenses for all of the benchmarks in
  under 30 seconds (\S\ref{sec:evaluation}). 
\end{itemize}
We close with a discussion of related work (\S\ref{sec:related}) and
concluding thoughts (\S\ref{sec:conc}).

@@

For each benchmark in our benchmark suite, we ran every mode on it with a
timeout of 30 seconds, and averaged the result over 5 runs.
Figure~\ref{fig:times} summarizes the results of these tests. We find that our
algorithm is able to synthesize all of the benchmarks in our benchmark suite in
under 30 seconds. Without compositional synthesis, the synthesis algorithm is
able to solve 26 problem instances. Recall from
Section~\ref{subsec:optimizations} that users may tune the termination metric.
Success on this set of benchmarks requires we tune that termination codition
(e.g. for some benchmarks, allowing the system flexibility to spend more time
considering more of the search space). In
Section~\ref{subsec:effect-of-heuristics}, we will consider the effect of fixing
the termination metric. In practice, we found customizing the termination metric
to be quite easy: if manual inspection of the lens showed there were too few
bijections, we would increase the termination parameter. If synthesis took too
long, we would decrease it.

@@

While our tool uses types and examples to specify invertable transformations, other
tools have been shown to synthesize a backwards transformation from ordinary code designed
to implement the forwards transformation.  For instance, Hu and
D'Antoni~\cite{program-inversion-symbolic-transducers} show how to invert transformations
using symbolic finite automata, and
Voigtl\"{a}nder~\cite{bidirectionalization-for-free} demonstrates how to
construct reverse transformations from forwards transformations
by exploiting parametricity properties.  These kinds of tools are very useful, but in different
circumstances from ours (namely, when one already has the code to implement one direction of
the transformation).

More broadly, information theory and probabilistic languages are common tools in
natural language understanding, machine translation, information retrieval, data
extraction and grammatical inference (see Pereira~\cite{pereira:info-theory},
for an introduction). Indeed, our work was inspired, in part, by the PADS format
inference tool~\cite{pads:synthesis}, which was in turn inspired by earlier work
on probabilistic gramatical inference and information
extraction~\cite{kushmerick-thesis,arasu:extracting}. PADS did not synthesize
data transformers, and we are not aware of the use of information-theoretic
principles in type-directed or syntax-guided synthesis of DSL programs.

>>>>

One convenient way to interconvert between these varied data formats is to
use a \emph{lens}~\cite{Focal2005}---a bidirectional program that can
transform data represented in a format $S$ to a format $T$, and vice versa,
offering ``round tripping'' guarantees that help ensure data is not lost or
corrupted as it is transformed back and forth between different
representations.

While domain-specific languages for writing lenses can facilitate building
principled data transformations, \textit{synthesizing} lenses 
can make the task even easier.  Indeed, lens languages offer excellent
targets for synthesis, since their specialized, sub-Turing-complete syntax
and expressive type systems drastically
reduce the space of possible programs, making search-based synthesis easier. 
%
Recent work~\cite{optician} has exploited this observation, demonstrating that
it is feasible to synthesize quite complex examples, albeit only for a simple
form of lenses---so-called bijective lenses. Specifically, given a pair of types
describing the intended source and target formats plus a small number of
examples of corresponding data, the bijective synthesis algorithm will find a
bijective lens that converts between data stored the provided formats. Often,
these algorithms can find an appropriate lens in less than a second without
requiring any examples.
%, with just types describing the source and target formats.
% \dpw{I would cite the ICFP paper (cite string:
%   maina+:quotient-synthesis) as to appear but I'm not sure how to do that
%   without possibly violating the spirit of double-blind reviewing. ie, if it
%   hasn't appeared yet, how could we know the content if we aren't the authors?
%   Mentioning it in the 3rd person does not seem to suffice ...} \ksf{The ICFP
%   2018 website lists the paper; I think it is okay to cite it. The reviewer can
%   assume we contacted the authors for a copy. We could ask Stephanie for a
%   ruling if we want to be sure. One of the principles of double blind as adopted
%   by SIGPLAN is not to reduce the quality of the paper; omitting this citation
%   would have that affect.}\bcp{+1}\saz{Stephanie says we should cite it in the
%   third person, even if that's a bit awkward, but otherwise citing it is OK.}
One reason that bijective lens synthesis is so effective, even relative to successful
synthesis projects in other domains, is that there are not very
many bijections between related data formats (unless the
same kind of data appears in the two formats many times). 
Hence, if the synthesis algorithm can
find any bijection at all, then it is very likely to find the one the
user wants.

However, there is a price for narrowing the synthesis domain so
dramatically: Many real data formats are almost, but not quite,
bijections. More often two related data formats have some data in
common, but also contain some differences. For instance, two related
tables might share four of five columns, but each has a different
fifth column. Alternately, one ad hoc configuration file might include
some file-specific meta data, such as a date, a time or a reference
number, whereas the other does not. Indeed of the 48 benchmark problems
described in Section \ref{sec:evaluation}, all of the benchmarks
taken from Flash Fill~\cite{flashfill} and many of the benchmarks that
synchronize between two ad hoc file formats have this 
characteristic, 13 benchmarks in total. Consequently, searching
only for strict bijections is often not applicable in practice.

@@

Initially, we hypothesized that extending the synthesis algorithms
of \citet{optician} to operate over symmetric rather than
bijective lenses
would be relatively straightforward:  Simply replace the bijective
combinators with symmetric ones 
and search using similar heuristics.  However, 
this na\"ive approach runs into two difficulties.

@@

In summary, our contributions are:\bcp{The section references do not proceed
  in order, so this doesn't work very well as an outline of the paper.
  Should we try to reorder the contributions?  Write a separate outline?
  Live with it as-is?}\afm{Personally, I don't see it as a problem.  It's nice
  when the contributions mirror the paper, I am not jarred when they don't.}
\begin{itemize}
\item We define {\em simple symmetric lenses}, with the
  the useful property that they do not maintain any hidden state
  (\S\ref{sec:overview}).  
  We present a collection of combinators for defining these lenses and
  describe how to 
  calculate their likelihoods (\S\ref{sec:ssl}). 
\item  We show that simple symmetric lenses
  fall between asymmetric lenses~\cite{Focal2005-long2} and full
  symmetric lenses~\cite{symmetric-lenses} in expressiveness
  (\S\ref{sec:related}).  We further show that the class of simple
  lenses can be characterized semantically as the subset of full
  symmetric lenses for which a simple ``forgetfulness'' property holds.
\item We show how to synthesize simple symmetric lenses (including all
  asymmetric lenses as a special case), using a novel application of stochastic
  regular expressions to guide the search process (\S\ref{sec:synthesis}). 
  To support that effort, we show that the star semi-ring equivalences on
  regular 
  expressions remain valid when applied to stochastic regular
  expressions, preserving the probability distributions as well as the languages
  (\S\ref{sec:sre}).
\item We extend the Boomerang implementation with simple symmetric lenses,
  demonstrating that they integrate smoothly with 
  other lens extensions such as quotient lenses~\cite{quotientlenses} and matching
  lenses~\cite{matchinglenses}.  We evaluate our
  implementation on a set of 48 lens synthesis benchmarks drawn from
  data cleaning, view maintenance, 
  and file synchronization tasks.  We demonstrate that we can
  synthesize simple symmetric lenses for all of the benchmarks in
  under 30 seconds (\S\ref{sec:evaluation}). 
\end{itemize}
We close with a discussion of related work (\S\ref{sec:related}) and
concluding thoughts (\S\ref{sec:conc}).

@@

For each benchmark in our benchmark suite, we ran every mode on it with a
timeout of 60 seconds, and averaged the result over 5 runs.
Figure~\ref{fig:times} summarizes the results of these tests. We find that our
algorithm is able to synthesize all of the benchmarks in our benchmark suite in
under 30 seconds. Without compositional synthesis, the synthesis algorithm is
able to solve 26 problem instances. Recall from
Section~\ref{subsec:optimizations} that users may tune the termination metric.
Success on this set of benchmarks requires we tune that termination codition
(e.g. for some benchmarks, allowing the system flexibility to spend more time
considering more of the search space). In
Section~\ref{subsec:effect-of-heuristics}, we will consider the effect of fixing
the termination metric. In practice, we found customizing the termination metric
to be quite easy: if manual inspection of the lens showed there were too few
bijections, we would increase the termination parameter. If synthesis took too
long, we would decrease it.

@@

While our tool uses types and examples to specify invertable transformations, other
tools have been shown to synthesize a backwards transformation from ordinary code designed
to implement the forwards transformation.  For instance, \citet{program-inversion-symbolic-transducers} show how to invert transformations
using symbolic finite automata, and
\citet{bidirectionalization-for-free} demonstrates how to
construct reverse transformations from forwards transformations
by exploiting parametricity properties.  These kinds of tools are very useful, but in different
circumstances from ours (namely, when one already has the code to implement one direction of
the transformation).

More broadly, information theory and probabilistic languages are common tools in
natural language understanding, machine translation, information retrieval, data
extraction and grammatical inference (see \citet{pereira:info-theory},
for an introduction). Indeed, our work was inspired, in part, by the PADS format
inference tool~\cite{pads:synthesis}, which was in turn inspired by earlier work
on probabilistic gramatical inference and information
extraction~\cite{kushmerick-thesis,arasu:extracting}. PADS did not synthesize
data transformers, and we are not aware of the use of information-theoretic
principles in type-directed or syntax-guided synthesis of DSL programs.
